### 2015-12-08 Irregular-occ/Irregular-occ.tex
.

\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton for 2015-2017
%\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
%\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
%\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
%\fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
%\fnmsep\thanks{\textbf{Partially based on observations made at Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
}

\author{ A. R. Gomes-Júnior
          \inst{1},
          M. Assafin 
          \inst{1} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
          L. Beauvalet
          \inst{2,3},
          R. Vieira-Martins
          \inst{1,2,4} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
%          J.-E. Arlot
%          \inst{4},      
          J. I. B. Camargo
          \inst{2,4},     
          B. E. Morgado
          \inst{1,2}
          F. Braga-Ribas
          \inst{2,5}, 
%          D. N. da Silva Neto
%          \inst{6},
%          A. H. Andrei
%          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
%          A. Dias-Oliveira
%          \inst{2},
%          G. Benedetti-Rossi
%          \inst{2},
%          Y. Duchemin
%          \inst{4,7},
%          J. Desmars
%          \inst{4},
%          V. Lainey
%          \inst{4},
%          W. Thuillot
%          \inst{4}
          }

@@

   \institute{Observatório do Valongo/UFRJ, Ladeira Pedro Antônio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
              \email{altair08@astro.ufrj.br}
              \and
Observatório Nacional/MCT, R. General José Cristino 77,
CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
              \email{rvm@on.br}
              \and
Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France
              \and
Laboratório Interinstitucional de e-Astronomia - LIneA, Rua Gal. José Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil
%              \and
%Institut de mécanique céleste et de calcul des éphémérides - Observatoire %de Paris, UMR 8028 du CNRS,
%77 Av. Denfert-Rochereau, 75014 Paris, France\\
%              \email{arlot@imcce.fr}
              \and
Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
%              \and
%Centro Universitário Estadual da Zona Oeste, Av. Manual Caldeira de Alvarenga 1203, CEP 23.070-200 Rio de Janeiro RJ, Brazil
%              \and
%              ESIGELEC-IRSEEM, Technopôle du Madrillet, Avenue Galilée, 76801 Saint-Etienne du Rouvray, France
              }

@@

\abstract
{Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system. It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin. The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.}
{We aim to predict stellar occultations for the eight biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.}
{We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument. These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations. We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future for the satellites of Jupiter. For Phoebe, Nereid and Triton we used the JPL ephemeris.}
{We managed to identify 687 candidates of stellar occultations between the period of January, 2015 and December, 2017. We made observational tests for the prediction of the event of Himalia in March 03, 2015. The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
{The comparisons between the predictions and the observation tests show a good agreement. We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.}

@@

No observation of a stellar occultation by an irregular satellite was published up to date. Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries. But, in fact, given their distances to us, current ephemeris and star positions are precise enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth. For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds). Thus, in this case, the cumulated uncertainty of ephemeris and star position must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.

\begin{table}
\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
\begin{centering}
\begin{tabular}{lccc}
\hline  \hline
\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & mas\tablefootmark{a}  & km & References \tabularnewline
\hline
Ananke & 8 & 29 & 1 \tabularnewline
Carme & 13 & 46 & 1 \tabularnewline
Elara & 24 & 86 & 1 \tabularnewline
Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
Leda & 5 & 20 & 1 \tabularnewline
Lysithea & 10 & 36 & 1 \tabularnewline
Pasiphae & 17 & 62 & 1 \tabularnewline
Sinope & 10 & 37 & 1 \tabularnewline
\hdashline
Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
\hdashline
Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
\hline
\end{tabular}
\tablebib{
(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
\end{centering}
\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
\tablefoottext{b}{From Cassini observation.}
\tablefoottext{c}{From Voyager-2 observation.}
\par
\end{table}

@@

In this work, the STE was used in all but three satellites, not Jovian ones. For these 3 cases, the JPL ephemeris was used instead. The JPL ephemeris for Phoebe is of enough high quality for the short term, mostly thanks to the observations by Cassini. In the case of Triton, the JPL ephemeris is also good enough, as the relatively large diameter of Triton favours the positive detection of occultations. A new short term ephemeris for Nereid is however still needed, but its production was beyond the scope of this work, so we used the JPL ephemeris in this case too.

@@

For Triton and Nereid, the candidates for stellar occultations in 2015 and 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}. This catalogue contains the stars in the path of Neptune in the sky up to mid-2016. The catalogue was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector. The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm.

A total of 687 events were identified between January 2015 and December 2017. In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig. \ref{Fig: ocultacao} as an example). Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS.

\begin{table}
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from 2015 up to 2017.}
\begin{centering}
\begin{tabular}{lcccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2015  & 2016 & 2017 & Total \tabularnewline
\hline
Ananke & 21 & 12 & 16 & 49 \tabularnewline
Carme & 26 & 20 & 14 & 60 \tabularnewline
Elara & 32 & 14 & 16 & 62 \tabularnewline
Himalia & 26 & 15 & 12 & 53 \tabularnewline
Leda & 13 & 8 & 24 & 45 \tabularnewline
Lysithea & 22 & 16 & 11 & 49 \tabularnewline
Pasiphae & 25 & 20 & 19 & 64 \tabularnewline
Sinope & 19 & 15 & 21 & 55 \tabularnewline
\hdashline
Phoebe\tablefootmark{a} & 13 & 32 & 98 & 143 \tabularnewline
\hdashline
Nereid\tablefootmark{a} & 41\tablefootmark{b} & 11\tablefootmark{b} & 1 & 53 \tabularnewline
Triton\tablefootmark{a} & 37\tablefootmark{b} & 16\tablefootmark{b} & 1 & 54 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
Occultations predicted using the UCAC4 catalogue and STE ephemeris.
\tablefoottext{a}{Using JPL ephemeris.}
\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
\end{table}

@@

\section{Occultation tests} \label{Sec: testes}

Observing a stellar occultation demands a great effort. And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites. So, before we start a large observational campaign, we tested some occultation predictions for large targets, to assess the quality of the predictions.

The tests consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other. Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}. The relative positions of the star and of the satellites were used to check the original predictions.

Two occultation tests were performed, one by Himalia that occurred on March 3, 2015 and the second by Elara that occurred on March 30, 2015. For each event, four situations were considered:
\begin{enumerate}
\item Prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were separated (different FOVs);
\item Same as 3 but with the star and the satellite close in the same FOV.
\end{enumerate}

Figure \ref{Fig: occ-Himalia} shows a zoomed in occultation map with the four situations for Himalia. The line in blue is center of the shadow for the situation 1 in the list above. The line in red using JPL ephemeris (situation 2). The line in black was made from obtained positions on February 22 observed with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m). On that day, Himalia and the star were observed in separate FOVs as they were still far apart (situation 3). On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event. Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4). From the calculated offsets, the center of the shadow was obtained. Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km from the shadow path). This was not necessary for testing the prediction.

@@

\begin{figure*}
\begin{centering}
\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
\caption{Predictions for Himalia: The straight lines show the center of the shadow over time. In blue is for the prediction using the STE ephemeris showed in Sec. \ref{Sec: integration}. In red the shadow using the JPL ephemeris. In black we apply offsets to the positions of star and satellite accordingly the observations made at February 22. In green, the same for the previous one but with observations made at March 03 when the objects were close to each other. The dots show the closest approach of the shadow for each prediction. The scales are relative to the closest approach of the prediction using the STE ephemeris. The bars show the estimated diameter of the satellite and estimated error of the prediction.}
\label{Fig: occ-Himalia}
\end{centering}
\end{figure*}

In this event, it is possible to see that the shade does not vary by much among the four predictions being smaller than the estimated diameter of the satellite and error. This suggests that there was a good probability of observing the event. In fact, the largest differences between the shadows of the four situations were 36s, in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows (see table \ref{Tab: comparison-Himalia}).

\begin{table}
\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.}
\begin{centering}
\begin{tabular}{lcc}
\hline  \hline
\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
Method  & Instant of C/A  & C/A   \tabularnewline
\hline
STE & 00:39:51 UTC & $0\farcs703$ \tabularnewline
JPL & -26s & +11mas (36km) \tabularnewline % (284km)
Feb. 22 Obs. & -14s & -20mas (65km) \tabularnewline % (153km)
Mar. 03 Obs. & -36s & -09mas (29km) \tabularnewline % (393km)
\hline
\end{tabular}
\par\end{centering}
%*Closest Approach
\end{table}

The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter. The event was predicted to occur at March 30, 2015. The observations were made on March 25 and April 2, 2015 with the Zeiss telescope. On the night of April 2 the star and satellite could still be observed in the same FOV. Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia. %Still, the differences between the maps obtained were relatively small. 
The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).

\begin{table}
\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
\begin{centering}
\begin{tabular}{lcc}
\hline  \hline
\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
Method  & Instant of C/A  & C/A  \tabularnewline
\hline
STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
JPL & +02s & +57mas(195km) \tabularnewline
Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
\hline
\end{tabular}
\par\end{centering}
\end{table}

@@

We presented stellar occultations for the period of 2015-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid. The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.

The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2015 and 2016 for Triton and Nereid. In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec. \ref{Sec: predictions}). From this, a total of 687 events are foreseen.

@@

The tests made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area. %Fig. \ref{Fig: occ-Himalia} shows the maps made for 4 different offset corrections as explained in Sec. \ref{Sec: testes}. Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
The results show satisfying small offsets. %We can expect similar results for other satellites and our method should allow successful stellar occultation.

>>>>

\title{Stellar occultation predictions for  10 irregular satellites of giant planets plus Triton up to 2017
%\fnmsep\thanks{Table 8 are only available in electronic form at the CDS via anonymous ftp to cdsarc.u-strasbg.fr (130.79.128.5) or via http://cdsweb.u-strasbg.fr/cgi-bin/qcat?J/A+A/ and IAU NSDC data base at www.imcce.fr/nsdc.}
%\fnmsep\thanks{The complete version of Table 8 is available through CDS and IAU NSDC data base at www.imcce.fr/nsdc.}
%\fnmsep\thanks{Partially based on observations made at Laborat\'orio Nacional de Astrof\'{\i}sica (LNA), Itajub\'a-MG, Brazil.}
%\fnmsep\thanks{Partially based on observations through the ESO runs 079.A-9202(A), 075.C-0154, 077.C-0283 and 079.C-0345.}
%\fnmsep\thanks{\textbf{Partially based on observations made at Observatoire de Haute Provence (OHP), F-04870 Saint-Michel l'observatoire, France}}
}

\author{ A. R. Gomes-Júnior
          \inst{1},
          M. Assafin 
          \inst{1} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
          L. Beauvalet
          \inst{2,3},
          R. Vieira-Martins
          \inst{1,2,4} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/IMCCE, 77 Avenue Denfert Rochereau 75014 Paris, France},
%          J.-E. Arlot
%          \inst{4},      
          J. I. B. Camargo
          \inst{2,4},     
          J. Desmars
          \inst{5},
          B. E. Morgado
          \inst{1,2},
          F. Braga-Ribas
          \inst{2,6}, 
%          D. N. da Silva Neto
%          \inst{6},
%          A. H. Andrei
%          \inst{1,2} \fnmsep\thanks{Affiliated researcher at Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau, 75014 Paris, France},
%          A. Dias-Oliveira
%          \inst{2},
%          G. Benedetti-Rossi
%          \inst{2},
%          Y. Duchemin
%          \inst{4,7},
%          V. Lainey
%          \inst{4},
%          W. Thuillot
%          \inst{4}
          }

@@

   \institute{Observatório do Valongo/UFRJ, Ladeira Pedro Antônio 43,
CEP 20.080-090 Rio de Janeiro - RJ, Brazil\\
              \email{altair08@astro.ufrj.br}
              \and
Observatório Nacional/MCT, R. General José Cristino 77,
CEP 20921-400 Rio de Janeiro - RJ, Brazil\\
              \email{rvm@on.br}
              \and
Observatoire de Paris/SYRTE, 77 Avenue Denfert Rochereau 75014 Paris, France
              \and
Laboratório Interinstitucional de e-Astronomia - LIneA, Rua Gal. José Cristino 77, Rio de Janeiro, RJ 20921-400, Brazil
              \and
Institut de mécanique céleste et de calcul des éphémérides - Observatoire de Paris, UMR 8028 du CNRS,
77 Av. Denfert-Rochereau, 75014 Paris, France
%              \email{arlot@imcce.fr}
              \and
Federal University of Technology - Paraná (UTFPR / DAFIS), Rua Sete de Setembro, 3165, CEP 80230-901, Curitiba, PR, Brazil
%              \and
%Centro Universitário Estadual da Zona Oeste, Av. Manual Caldeira de Alvarenga 1203, CEP 23.070-200 Rio de Janeiro RJ, Brazil
%              \and
%              ESIGELEC-IRSEEM, Technopôle du Madrillet, Avenue Galilée, 76801 Saint-Etienne du Rouvray, France
              }

@@

\abstract
{Due to their orbital configurations, it is common belief that the irregular satellites were probably captured by the giant planets in the early solar system. It is important to know their physical parameters, such as size, shape, albedo and composition, to trace back their true origin. The best ground-based technique to determine size and shape, and thus constrain the albedo and in a broader sense composition, is the observation of stellar occultations by these objects.}
{We aim to predict stellar occultations for the eight biggest irregular satellites of Jupiter: Himalia, Elara, Pasiphae, Carme, Lysithea, Sinope, Ananke and Leda, and for the irregular satellites Phoebe of Saturn and Nereid of Neptune, and also for Triton.}
{We identified candidates to stellar occultations by the irregular satellites from the UCAC4 catalogue and from a catalogue of stars in the sky-path of Neptune obtained from observations made with the ESO2p2/WFI (2.2 m Max-Planck ESO telescope with the Wide Field Imager) instrument. These catalogues were crossed with the ephemeris of the satellites to identify stellar occultations. We used a new ephemeris based solely on the observations from \cite{GomesJunior2015} to generate predictions for the short-time future for the satellites of Jupiter. For Phoebe, Nereid and Triton we used the JPL ephemeris.}
{We managed to identify 396 candidates of stellar occultations between the period of January, 2016 and December, 2017. We made observational tests for the prediction of the event of Himalia in March 03, 2015. The stars and objects were observed close to the date predicted and in the same field to minimize errors and the obtained relative satellite-star positions were used to evaluate the predictions.}
{The comparisons between the predictions and the observation tests show a good agreement. We discuss how the successful observation of a stellar occultation by these objects is quite possible and present some of those potential occultations.}

@@

No observation of a stellar occultation by an irregular satellite was published up to date, excluding Triton. Since their estimated sizes are very small (see Table \ref{Tab: satellite-diameter}), this may have discouraged earlier tries. But, in fact, given their distances to us, current ephemeris and star positions are precise enough for the prediction of the exact location and instant where the shadow of the occultation will cross the Earth. For instance, Himalia, supposedly the largest irregular satellite of Jupiter has an estimated size of 150 km \citep{Porco2003}, which is equivalent to an apparent size of about 40 mas (milliarcseconds). Thus, in this case, the cumulated uncertainty of ephemeris and star position must be around 40 mas to get good chances of observing a stellar occultation, which is quite feasible today.

\begin{table}
\caption{\label{Tab: satellite-diameter} Estimated diameter of the satellites and correspondent apparent diameter}
\begin{centering}
\begin{tabular}{lccc}
\hline  \hline
\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & mas\tablefootmark{a}  & km & Ref. \tabularnewline
\hline
Ananke & 8 & 29 & 1 \tabularnewline
Carme & 13 & 46 & 1 \tabularnewline
Elara & 24 & 86 & 1 \tabularnewline
Himalia & 41 & $(150\times120) \pm 20$\tablefootmark{b} & 2 \tabularnewline
Leda & 5 & 20 & 1 \tabularnewline
Lysithea & 10 & 36 & 1 \tabularnewline
Pasiphae & 17 & 62 & 1 \tabularnewline
Sinope & 10 & 37 & 1 \tabularnewline
\hdashline
Phoebe & 32 & $212 \pm 1.4$\tablefootmark{b} & 3 \tabularnewline
\hdashline
Nereid & 15 & $340 \pm 50$\tablefootmark{c} & 4 \tabularnewline
Triton & 124 & $2707 \pm 2.0$\tablefootmark{c} & 5 \tabularnewline
\hline
\end{tabular}
\tablebib{
(1) \cite{Rettig2001}; (2) \cite{Porco2003}; (3) \cite{Thomas2010}; (4) \cite{Thomas1991}; (5) \cite{Thomas2000}.}
\end{centering}
\tablefoottext{a}{Using a mean distance from Jupiter of 5 AU, from Saturn of 9 AU and from Neptune of 30 AU.}
\tablefoottext{b}{From Cassini observation.}
\tablefoottext{c}{From Voyager-2 observation.}
\par
\end{table}

@@

In this work, the STE was used in all but three satellites, not Jovian ones. For these 3 cases, the JPL ephemeris was used instead. The JPL ephemeris for Phoebe is of enough high quality for the short term, mostly thanks to the observations by Cassini. In the case of Triton, the JPL ephemeris is also good enough, as the relatively large diameter of Triton favours the positive detection of occultations. A new short term ephemeris for Nereid is however still needed, but its production was beyond the scope of this work due to its more complicated dynamic, so we used the JPL ephemeris in this case too.

@@

For Triton and Nereid, the candidates for stellar occultations in 2016 was searched using the WFI catalogue in the same way as the predictions for Centaurs and TNOs occultations by \cite{Assafin2010, Assafin2012} and \cite{Camargo2014}. This catalogue contains the stars in the path of Neptune in the sky up to mid-2016. The catalogue was generated by observations made at the ESO 2p2 telescope (IAU code 809) using the Wide Field Imager (WFI) CCD mosaic detector. The filter used was the broad-band R filter ESO\#844 with $\lambda_c$ = 651.725 nm and $\Delta\lambda$ = 162.184 nm.

A total of 396 events were identified between January 2016 and December 2017. In table \ref{Tab: satellite-occultation} we present the number of stellar occultations predicted by year for each satellite. Table \ref{Tab: occ-list} shows a sample of the catalogue of occultations generated and their parameters, which are necessary to produce occultation maps (see Fig. \ref{Fig: ocultacao} as an example). Since these objects are very small, the duration of each event is a few seconds. All the occultation tables and maps will be publicly available at the CDS. No star brighter than MagR*=18 was identified to be occulted by Triton in 2016. Since the flux drop during the such an occultation is very small, we don't publish this events.

\begin{table}
\caption{\label{Tab: satellite-occultation} Number of stellar occultations for each satellite from January, 2016 up to December, 2017.}
\begin{centering}
\begin{tabular}{lccc}
\hline  \hline
%\multicolumn{4}{c}{Diameter of the satellites} \tabularnewline
Satellite  & 2016 & 2017 & Total \tabularnewline
\hline
Ananke & 12 & 16 & 28 \tabularnewline
Carme & 20 & 14 & 34 \tabularnewline
Elara & 14 & 16 & 30 \tabularnewline
Himalia & 15 & 12 & 27 \tabularnewline
Leda & 8 & 24 & 32 \tabularnewline
Lysithea & 16 & 11 & 27 \tabularnewline
Pasiphae & 20 & 19 & 39 \tabularnewline
Sinope & 15 & 21 & 36 \tabularnewline
\hdashline
Phoebe\tablefootmark{a} & 32 & 98 & 130 \tabularnewline
\hdashline
Nereid\tablefootmark{a} & 11\tablefootmark{b} & 1 & 12 \tabularnewline
Triton\tablefootmark{a} & -- & 1 & 1 \tabularnewline
\hline
\end{tabular}
\par \end{centering}
Occultations predicted using the UCAC4 catalogue and STE ephemeris.
\tablefoottext{a}{Using JPL ephemeris.}
\tablefoottext{b} {Using the WFI catalogue as explained in Sec. \ref{Sec: predictions}}.
\end{table}

@@

\subsection{Occultation test} \label{Sec: testes}

Observing a stellar occultation demands a great effort. And, in our case, the shadow covers a very restricted area on Earth because of the size of the irregular satellites. Since no stellar occultation by irregular satellite was observed up to date, with the exception of Triton, and before we start a large observational campaign, we tested an occultation prediction for a large target, to assess the quality of the prediction.

The test consisted in observing the object and star to be occulted near the date of the event predicted when the two objects were present in the same field of view (FOV), preferably when the objects were close to each other. Thus, the relative positions between the two objects had minimal influence of the errors of the reference catalogue of stars used and possible field distortions \citep[and references therein]{Peng2008}. The relative positions of the star and of the satellites were used to check the original predictions.

The occultation tested was performed by Himalia that occurred on March 3, 2015. For the event, four situations were considered:
\begin{enumerate}[I]
\item Prediction with the STE ephemeris (see Sec. \ref{Sec: integration}), and the nominal UCAC4 position of the star;
\item Prediction with the JPL ephemeris and the nominal UCAC4 position of the star;
\item From star and satellite offsets calculated from observations made a few days before the occultation when the objects were separated (different FOVs);
\item Same as 3 but with the star and the satellite close in the same FOV.
\end{enumerate}

Table \ref{Tab: comparison-Himalia} shows the differences between the predictions in the four situations. For situation 3 we observed the objects on February 22 with the Zeiss telescope (diameter = 0.6m; FOV = $12\farcm6$; pixel scale = $0\farcs37 / px$) at the Observatório do Pico dos Dias (OPD, IAU code 874, 45\degr 34\arcmin 57\arcsec W, 22\degr 32\arcmin 04\arcsec S, 1864m). On that day, Himalia and the star were observed in separate FOVs as they were still far apart. On the night of the event, March 3, the objects were observed with Perkin-Elmer telescope (diameter = 1.6m; FOV = $5\farcm8$; pixel scale = $0\farcs17 / px$) at OPD just over an hour after the time scheduled for the event. Satellite and star were separated by about 16 arcsec, so very close to each other (situation 4). From the calculated offsets, the center of the shadow was obtained. Notice that the shadow path was not predicted to cross the OPD (which was located at almost 2000 km from the shadow path). This was not necessary for testing the prediction.

@@

%\begin{figure*}
%\begin{centering}
%\includegraphics[width=17.6cm]{figures/HIMALIA.eps}
%\caption{Predictions for Himalia: The straight lines show the center of the shadow over time. In blue is for the prediction using the STE ephemeris showed in Sec. \ref{Sec: integration}. In red the shadow using the JPL ephemeris. In black we apply offsets to the positions of star and satellite accordingly the observations made at February 22. In green, the same for the previous one but with observations made at March 03 when the objects were close to each other. The dots show the closest approach of the shadow for each prediction. The scales are relative to the closest approach of the prediction using the STE ephemeris. The bars show the estimated diameter of the satellite and estimated error of the prediction.}
%\label{Fig: occ-Himalia}
%\end{centering}
%\end{figure*}

In this event, it is possible to see that the shade does not vary by much among the four predictions being smaller than the estimated diameter of the satellite and error. This suggests that there was a good probability of observing the event. In fact, the largest differences between the shadows of the four situations were 36s in time along the shadow path and 101km (31 mas) in the direction perpendicular to the shadows.

\begin{table}
\caption{\label{Tab: comparison-Himalia} Comparison between the predictions of the Himalia occultation at March 03, 2015.}
\begin{centering}
\begin{tabular}{lccc}
\hline  \hline
\multicolumn{4}{c}{Difference from STE prediction} \tabularnewline
Method  & Instant of C/A  & C/A & Sit.   \tabularnewline
\hline
STE & 00:39:51 UTC & $0\farcs703$ & I \tabularnewline
JPL & -26s & +11mas (36km) & II \tabularnewline % (284km)
Feb. 22 Obs. & -14s & -20mas (65km) & III \tabularnewline % (153km)
Mar. 03 Obs. & -36s & -09mas (29km) & IV \tabularnewline % (393km)
\hline
\end{tabular}
\par\end{centering}
C/A: geocentric closest approach; Sit: Situation test considered.
\end{table}

%The second test was with the satellite Elara, which is the second largest irregular satellite of Jupiter. The event was predicted to occur at March 30, 2015. The observations were made on March 25 and April 2, 2015 with the Zeiss telescope. On the night of April 2 the star and satellite could still be observed in the same FOV. Due to Elara being fainter than Himalia by 2 magnitudes, dispersions of the satellite positions on both nights ended up being higher than for Himalia. %Still, the differences between the maps obtained were relatively small. 
%The largest differences between them were 144s (186 mas) in time along the shadow path and 525 km (153 mas) perpendicular to it (see table \ref{Tab: comparison-Elara}).

%\begin{table}
%\caption{\label{Tab: comparison-Elara} Comparison between the predictions of Elara occultation at March 30, 2015.}
%\begin{centering}
%\begin{tabular}{lcc}
%\hline  \hline
%\multicolumn{3}{c}{Difference from STE prediction} \tabularnewline
%Method  & Instant of C/A  & C/A  \tabularnewline
%\hline
%STE & 01:45:13 UTC & $1\farcs082$ \tabularnewline
%JPL & +02s & +57mas(195km) \tabularnewline
%Mar. 25 Obs. & -70s & +20mas (68km) \tabularnewline
%Apr. 02 Obs. & +74s & -96mas (330km) \tabularnewline
%\hline
%\end{tabular}
%\par\end{centering}
%\end{table}

@@

We presented stellar occultations for the period of 2016-2017 for eight irregular satellites of Jupiter: Ananke, Carme, Elara, Himalia, Leda, Lysithea, Pasiphae, and Sinope; one satellite of Saturn: Phoebe; and two satellites of Neptune: Triton and Nereid. The procedure used was the same as that for the prediction of stellar occultations by Pluto and its satellites in \cite{Assafin2010} and by Centaurs and TNOs in \cite{Assafin2012} and \cite{Camargo2014}.

The candidate stars were searched in the UCAC4 catalogue, except for the candidates in 2016 for Triton and Nereid. In this case, we used the WFI catalogue that was generated from observations made with ESO2p2/WFI CCD mosaic that covered the path of Neptune in the sky-plane up to 2016 (see Sec. \ref{Sec: predictions}). From this, a total of 396 events are foreseen.

@@

The test made with an occultation expected to happen in March 03, 2015 for Himalia showed that this event would probably have been observed successfully in case there were observers available in the shadow area. %Fig. \ref{Fig: occ-Himalia} shows the maps made for 4 different offset corrections as explained in Sec. \ref{Sec: testes}. Table \ref{Tab: comparison-Himalia} shows numerically the difference between the maps.
%A similar test was made for an occultation of Elara in March 30, 2015. %The difference between the maps in this case are listed in Table \ref{Tab: comparison-Elara}.
The results show satisfying small offsets. %We can expect similar results for other satellites and our method should allow successful stellar occultation.
